---
title: "Plots and Graphs, Minas"
output: html_document
date: "2026-07-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Codes used to create plots and figures**

*Graphs based on Table 2. Direct cancer expenditure in 18 EECCA countries in 2024.*

```{r}
# Armenia with the same methodlogy as others
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(patchwork)

# Data: Country, Per-Capita ($PPP), Per-Incident-Case ($PPP)

rank_raw <- tibble::tribble(
  ~Country,        ~Region,          ~PerCapita, ~PerIncident,
  "Armenia",       "Caucasus",       145,        41070,
  "Azerbaijan",    "Caucasus",       27,         18334,
  "Georgia",       "Caucasus",       211,        48535,
  "Afghanistan",   "Central Asia",   0.04,       67,
  "Kazakhstan",    "Central Asia",   80.6,       55586,
  "Kyrgyzstan",    "Central Asia",   6.2,        5003,
  "Tajikistan",    "Central Asia",   17.5,       22506,
  "Turkmenistan",  "Central Asia",   52.8,       51626,
  "Uzbekistan",    "Central Asia",   3.7,        3523,
  "Belarus",       "Eastern Europe", 101,        19839,
  "Estonia",       "Eastern Europe", 463,        75978,
  "Latvia",        "Eastern Europe", 294,        48530,
  "Lithuania",     "Eastern Europe", 407,        71783,
  "Moldova",       "Eastern Europe", 43.9,       9290,
  "Russia",        "Eastern Europe", 129,        29152,
  "Slovakia",      "Eastern Europe", 236,        40793,
  "Slovenia",      "Eastern Europe", 452,        67459,
  "Ukraine",       "Eastern Europe", 30.7,       7554
)

# colourblind-safe region palette (Okabe-Ito)
rank_pal <- c("Caucasus"       = "#E69F00",
              "Central Asia"   = "#009E73",
              "Eastern Europe" = "#0072B2")

#Panel 1: Per-capita, ranked high -> low
rank_pc <- rank_raw %>% mutate(Country = reorder(Country, PerCapita))

p_pc <- ggplot(rank_pc, aes(x = PerCapita, y = Country, fill = Region)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = comma(PerCapita, accuracy = 0.1)),
            hjust = -0.15, size = 2.8, family = "serif") +
  scale_fill_manual(values = rank_pal, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Per-capita cancer expenditure",
       x = "Per capita ($PPP)", y = NULL) +
  theme_classic(base_size = 11, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 12),
        legend.position = "none")

#Panel 2: Per-incident-case, ranked high -> low 
rank_pi <- rank_raw %>% mutate(Country = reorder(Country, PerIncident))

p_pi <- ggplot(rank_pi, aes(x = PerIncident, y = Country, fill = Region)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = comma(PerIncident)),
            hjust = -0.15, size = 2.8, family = "serif") +
  scale_fill_manual(values = rank_pal, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18)), labels = comma) +
  labs(title = "Per-incident-case cancer expenditure",
       x = "Per incident case ($PPP)", y = NULL) +
  theme_classic(base_size = 11, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 12),
        legend.position = "none")

#Combine with a shared bottom legend
rank_combo <- (p_pc | p_pi) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Cancer Expenditure Rankings Across 18 EECCA Countries",
    theme = theme(plot.title = element_text(hjust = .5, size = 13, family = "serif"))
  ) &
  theme(legend.position = "bottom")

print(rank_combo)

#ggsave("cancer_expenditure_rankings_main.png", rank_combo,
#       width = 12, height = 6.5, dpi = 400, bg = "white")

```

``` {r}
# SENSITIVITY: Cancer expenditure ranking — per capita & per incident case
# Armenia on national NHA basis (90.8% OOP): per-capita 316, per-incident 89,418
#above figures on Armenia is on the GHED 80% OOP basis. For other countries no NHA data is available. 

#Data: Country, Per-Capita ($PPP), Per-Incident-Case ($PPP)
rank_raw_sens <- tibble::tribble(
  ~Country,        ~Region,          ~PerCapita, ~PerIncident,
  "Armenia",       "Caucasus",       316,        89418,   # national NHA basis
  "Azerbaijan",    "Caucasus",       27,         18334,
  "Georgia",       "Caucasus",       211,        48535,
  "Afghanistan",   "Central Asia",   0.04,       67,
  "Kazakhstan",    "Central Asia",   80.6,       55586,
  "Kyrgyzstan",    "Central Asia",   6.2,        5003,
  "Tajikistan",    "Central Asia",   17.5,       22506,
  "Turkmenistan",  "Central Asia",   52.8,       51626,
  "Uzbekistan",    "Central Asia",   3.7,        3523,
  "Belarus",       "Eastern Europe", 101,        19839,
  "Estonia",       "Eastern Europe", 463,        75978,
  "Latvia",        "Eastern Europe", 294,        48530,
  "Lithuania",     "Eastern Europe", 407,        71783,
  "Moldova",       "Eastern Europe", 43.9,       9290,
  "Russia",        "Eastern Europe", 129,        29152,
  "Slovakia",      "Eastern Europe", 236,        40793,
  "Slovenia",      "Eastern Europe", 452,        67459,
  "Ukraine",       "Eastern Europe", 30.7,       7554
)

# colourblind-safe region palette (Okabe-Ito)
rank_pal_sens <- c("Caucasus"       = "#E69F00",
                   "Central Asia"   = "#009E73",
                   "Eastern Europe" = "#0072B2")

#Panel 1: Per-capita, ranked high -> low
rank_pc_sens <- rank_raw_sens %>% mutate(Country = reorder(Country, PerCapita))

p_pc_sens <- ggplot(rank_pc_sens, aes(x = PerCapita, y = Country, fill = Region)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = comma(PerCapita, accuracy = 0.1)),
            hjust = -0.15, size = 2.8, family = "serif") +
  scale_fill_manual(values = rank_pal_sens, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Per-capita cancer expenditure",
       x = "Per capita ($PPP)", y = NULL) +
  theme_classic(base_size = 11, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 12),
        legend.position = "none")

#Panel 2: Per-incident-case, ranked high -> low
rank_pi_sens <- rank_raw_sens %>% mutate(Country = reorder(Country, PerIncident))

p_pi_sens <- ggplot(rank_pi_sens, aes(x = PerIncident, y = Country, fill = Region)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = comma(PerIncident)),
            hjust = -0.15, size = 2.8, family = "serif") +
  scale_fill_manual(values = rank_pal_sens, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18)), labels = comma) +
  labs(title = "Per-incident-case cancer expenditure",
       x = "Per incident case ($PPP)", y = NULL) +
  theme_classic(base_size = 11, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 12),
        legend.position = "none")

#Combine with a shared bottom legend 
rank_combo_sens <- (p_pc_sens | p_pi_sens) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Cancer Expenditure Rankings Across 18 EECCA Countries (Sensitivity: Armenia NHA basis)",
    theme = theme(plot.title = element_text(hjust = .5, size = 13, family = "serif"))
  ) &
  theme(legend.position = "bottom")

print(rank_combo_sens)

#ggsave("cancer_expenditure_rankings_sensitivity.png", rank_combo_sens,
#       width = 12, height = 6.5, dpi = 400, bg = "white")
```


*Figure 2. Per-componnet expenditure*
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(patchwork)   



bar_raw <- tibble::tribble(
  ~Country,     ~Screening,   ~Pharmaceuticals, ~Other,
  "Lithuania",  21.7,         200,              182.8,
  "Slovenia",   26840103.33,  84322620,         360593845.9,
  "Moldova",    4790500,      0,                552520230,
  "Georgia",    NA,           151011920.6,      133135314.9,
  "Estonia",    7511966,      42146735,         243356243,
  "Kazakhstan", NA,           56300788282,      137637211718,
  "Latvia",     5789797,      74935311.68,      95005708,
  "Armenia",    83,           14413.70,         105918
)

bar_long <- bar_raw %>%
  pivot_longer(-Country, names_to = "Category", values_to = "Value") %>%
  group_by(Country) %>%
  mutate(pct = Value / sum(Value, na.rm = TRUE)) %>%  
  ungroup()

# order countries by screening share (NA -> treated as lowest, sorted to the end)
bar_order <- bar_long %>%
  filter(Category == "Screening") %>%
  arrange(desc(pct)) %>% pull(Country)   # NAs sink to the end automatically

bar_long <- bar_long %>%
  mutate(
    Country  = factor(Country, levels = bar_order),
    Category = factor(Category,
                      levels = c("Screening", "Pharmaceuticals", "Other"),
                      labels = c("Screening", "Pharmaceuticals", "Other medical services"))
  )

# colourblind-safe palette
bar_pal <- c(
  "Screening"              = "#E69F00",  # orange
  "Pharmaceuticals"        = "#0072B2",  # blue
  "Other medical services" = "#009E73"   # green
)

#Panel 1: full 100% composition
bar_main <- ggplot(bar_long, aes(x = Country, y = pct, fill = Category)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.3,
           position = position_stack(reverse = TRUE)) +
  geom_text(aes(label = ifelse(!is.na(pct) & pct >= 0.04,
                               paste0(round(pct*100), "%"), "")),
            position = position_stack(vjust = 0.5, reverse = TRUE),
            colour = "white", size = 3.5, family = "serif") +
  scale_fill_manual(values = bar_pal, name = NULL) +
  scale_y_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Composition of GCHCFS Expenditure on Cancer", x = NULL,
       y = "% of expenditure") +
  theme_classic(base_size = 12, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 13),
        legend.position = "bottom")

#Panel 2: screening share only (countries with data only) 
bar_scr <- bar_long %>%
  filter(Category == "Screening", !is.na(pct)) %>%   # drop Georgia & Kazakhstan
  mutate(Country = droplevels(Country))              # remove their empty x slots

bar_zoom <- ggplot(bar_scr, aes(x = Country, y = pct)) +
  geom_col(width = 0.72, fill = bar_pal["Screening"]) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%")),
            vjust = -0.4, size = 3.5, family = "serif") +
  scale_y_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Screening Share, Zoomed-in", x = NULL, y = "%screening share") +
  theme_classic(base_size = 12, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 12))

# combine
bar_combo <- bar_main / bar_zoom + plot_layout(heights = c(2.4, 1))
print(bar_combo)

#ggsave("cancer_composition_with_detail.png", bar_combo,
#       width = 10, height = 8, dpi = 400, bg = "white")
```


*Figure 3. Longitudinal Analysis.*

```{r percapita-plot}

##all countries on one axis

pc7_data <- data.frame(
  Year       = c(2019, 2020, 2021, 2022, 2023, 2024, 2025),
  Armenia    = c(        NA,         NA, 14.53935393, 29.56563240, 31.16121038, 29.01402682,         NA),
  Estonia    = c(        NA, 321.9224796, 326.5245939, 295.4968819, 328.8153358, 365.7751777, 382.4455955),
  Kazakhstan = c(27.54365164, 33.58047086, 36.66758922, 44.53241539, 60.56892839, 58.01531222, 56.82907644),
  Moldova    = c(        NA, 22.84499350, 31.72142107, 37.12471370, 31.52723154, 46.73509602,         NA),
  Russia     = c(78.89591874, 106.0095318, 102.0952648, 92.09236764, 97.82014943, 92.58843846, 99.35289914),
  Slovakia   = c(        NA, 177.7578003, 202.1497472, 204.0481109, 195.1411889,         NA,         NA),
  Ukraine    = c(        NA, 14.40647411, 21.27716756, 19.02266181, 17.19708425, 16.86658272, 14.35476985)
)

pc7_long <- pc7_data %>%
  pivot_longer(-Year, names_to = "Country", values_to = "Value") %>%
  filter(!is.na(Value))

pc7_pal <- c(
  "Armenia"    = "#E69F00",
  "Estonia"    = "#56B4E9",
  "Kazakhstan" = "#009E73",
  "Moldova"    = "#F0E442",
  "Russia"     = "#0072B2",
  "Slovakia"   = "#D55E00",
  "Ukraine"    = "#CC79A7"
)

pc7_plot <- ggplot(pc7_long, aes(x = Year, y = Value,
                                 colour = Country, shape = Country)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2, fill = NA, stroke = 0.8) +
  scale_colour_manual(values = pc7_pal) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 7, 3)) +
  scale_x_continuous(breaks = seq(2019, 2025, 1)) +
  scale_y_log10(
    breaks = c(10, 20, 50, 100, 200, 400),
    labels = c("10", "20", "50", "100", "200", "400")
  ) +
  labs(
    title  = "Expenditure on cancer over the years",
    x = "Year",
    y = "Per capita expenditure on cancer (PPP$, log scale)",
    colour = "Country", shape = "Country"
  ) +
  theme_bw(base_size = 12, base_family = "serif") +
  theme(
    plot.title        = element_text(hjust = 0.5, size = 13),
    panel.grid.major  = element_line(colour = "grey85", linewidth = 0.3),  # subtle grid
    panel.grid.minor  = element_line(colour = "grey92", linewidth = 0.2),  # fainter minor grid
    panel.border      = element_rect(colour = "black", linewidth = 0.6, fill = NA),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(-1.5, "pt"),
    legend.position   = "right",                    # legend OUTSIDE, to the right
    legend.title      = element_text(face = "bold"),
    legend.key        = element_blank()
  )

print(pc7_plot)

#ggsave("percapita_plot_all.png", pc7_plot,
#       width = 9, height = 5.5, dpi = 400, bg = "white")
```

Now we make a plot for the appendix for each country. 

```{r percapita-facets}
pc7_facet_plot <- ggplot(pc7_long, aes(x = Year, y = Value, colour = Country)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8, fill = NA, stroke = 0.8) +
  scale_colour_manual(values = pc7_pal, guide = "none") +   # colour kept, legend off
  scale_x_continuous(breaks = seq(2019, 2025, 2)) +         # every 2 yrs (panels are small)
  scale_y_log10(
    breaks = c(10, 20, 50, 100, 200, 400),
    labels = c("10", "20", "50", "100", "200", "400")
  ) +
  facet_wrap(~Country, ncol = 4) +          # 7 panels, fixed (shared) axes
  labs(
    title = "Expenditure on cancer over the years",
    x = "Year",
    y = "Per capita expenditure on cancer (PPP$, log scale)"
  ) +
  theme_bw(base_size = 11, base_family = "serif") +
  theme(
    plot.title        = element_text(hjust = 0.5, size = 13),
    strip.background  = element_rect(fill = "grey85", colour = "black", linewidth = 0.4),
    strip.text        = element_text(face = "bold"),
    panel.grid.major  = element_line(colour = "grey85", linewidth = 0.3),
    panel.grid.minor  = element_line(colour = "grey92", linewidth = 0.2),
    panel.border      = element_rect(colour = "black", linewidth = 0.6, fill = NA),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(-1.5, "pt"),
    panel.spacing.x   = unit(1.2, "lines")   # <- add this
  )

print(pc7_facet_plot)

#ggsave("percapita_indiv.png", pc7_facet_plot,
#       width = 11, height = 6, dpi = 400, bg = "white")
```
in case questions, we also prepared the unlogged scale, where all figures are on the same plot. It is for the main text: 

```{r}
#linear unscaled individual
pc7_facet_linear1 <- ggplot(pc7_long, aes(x = Year, y = Value, colour = Country)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8, fill = NA, stroke = 0.8) +
  scale_colour_manual(values = pc7_pal, guide = "none") +
  scale_x_continuous(breaks = seq(2019, 2025, 2)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.10))) +
  facet_wrap(~Country, ncol = 4, scales = "free_y") +   # each panel its own y-axis
  labs(
    title = "Expenditure on cancer over the years",
    x = "Year",
    y = "Per capita expenditure on cancer (PPP$)"
  ) +
  theme_bw(base_size = 11, base_family = "serif") +
  theme(
    plot.title        = element_text(hjust = 0.5, size = 13),
    strip.background  = element_rect(fill = "grey85", colour = "black", linewidth = 0.4),
    strip.text        = element_text(face = "bold"),
    panel.grid.major  = element_line(colour = "grey85", linewidth = 0.3),
    panel.grid.minor  = element_line(colour = "grey92", linewidth = 0.2),
    panel.border      = element_rect(colour = "black", linewidth = 0.6, fill = NA),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(-1.5, "pt"),
    panel.spacing.x   = unit(1.2, "lines")
  )

print(pc7_facet_linear1)

#ggsave("percapita_indiv_linear_append.png", pc7_facet_linear1,
#       width = 13, height = 6, dpi = 400, bg = "white")
```


*Figure 4. Slovakia Per-Diagnosis Expenditure*

```{r}
#Slovakia — cancer expenditure by diagnosis
#MAIN: heatmap of cumulative % change vs 2020 baseline


library(scales)
library(forcats)

#Data 
slk_wide <- data.frame(
  Group = c("Breast (C50)", "Prostate (C61)", "Lung (C33-34)",
            "Colorectal (C18-20)", "Digestive (C15-17,C21-26)",
            "Female reproductive (C51-58)",
            "Male reproductive excl prostate (C60-63)", "Urinary (C64-68)",
            "Brain & CNS (C69-72)", "Endocrine (C73-75)", "Skin (C43-44)",
            "Hematological (C81-96)", "Other cancers"),
  Y2020 = c(72801194,47138623,33818367,64706117,46936834,19871601,6910067,
            36830870,12159130,5523525,31928288,81594888,61065648),
  Y2021 = c(85786218,55611553,35281189,72537910,49135855,23509555,4727888,
            39485599,11731517,6661874,39485263,104930247,65072312),
  Y2022 = c(84863967,57795744,34742289,67328057,46166641,21893158,4382376,
            34855076,10100063,6370295,34965251,111645400,61499294),
  Y2023 = c(101530746,57268984,36746939,50821205,29315873,13825655,2761234,
            24770670,5904579,3814457,45404893,112820946,54107953)
)


slk_long <- slk_wide %>%
  pivot_longer(starts_with("Y"), names_to = "Year", values_to = "Expenditure") %>%
  mutate(Year = as.numeric(gsub("Y", "", Year)))

# ---- Cumulative % change vs 2020 baseline
slk_change <- slk_long %>%
  group_by(Group) %>%
  mutate(
    PctChange = (Expenditure / Expenditure[Year == 2020] - 1) * 100,
    OrderKey  = PctChange[Year == 2023]      
  ) %>%
  ungroup() %>%
  mutate(Group = fct_reorder(Group, OrderKey, .desc = FALSE))


slk_lim <- max(abs(slk_change$PctChange))

#Heatmap 
slk_heat_change <- ggplot(slk_change,
                          aes(x = factor(Year), y = Group, fill = PctChange)) +
  geom_tile(colour = "grey96", linewidth = 0.5) +
  geom_tile(data = subset(slk_change, Year == 2020),
            fill = "grey90", colour = "grey96", linewidth = 0.5) +
  geom_text(aes(label = ifelse(Year == 2020, "ref",
                               paste0(ifelse(PctChange >= 0, "+", ""),
                                      round(PctChange), "%"))),
            size = 3.4, family = "serif",
            colour = ifelse(abs(slk_change$PctChange) > 45, "white", "grey15")) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
    limits = c(-slk_lim, slk_lim),
    labels = function(x) paste0(x, "%")
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(title = "Change in Cancer Expenditure by Diagnosis, Slovakia",
       x = NULL, y = NULL, fill = "Cum.change\nvs 2020") +
  theme_minimal(base_size = 11, base_family = "serif") +
  theme(
    plot.title   = element_text(hjust = 0.5, size = 13, face = "bold",
                                margin = margin(b = 10)),
    axis.text.y  = element_text(size = 10),
    axis.text.x  = element_text(size = 9),
    panel.grid   = element_blank(),
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8),
    plot.margin  = margin(t = 8, r = 8, b = 8, l = 8)
  )

print(slk_heat_change)

#ggsave("slovakia_change_heatmap_MAIN.png", slk_heat_change,
#       width = 8, height = 7, dpi = 400, bg = "white")
```

```{r}

# APPENDIX: stacked area of absolute composition

slk_area_data <- slk_long %>%
  mutate(Group = fct_reorder(Group, Expenditure, .fun = sum))

slk_area <- ggplot(slk_area_data, aes(x = Year, y = Expenditure, fill = Group)) +
  geom_area(alpha = 0.9, linewidth = 0) +
  scale_y_continuous(labels = label_number(scale = 1e-6)) +
  scale_x_continuous(breaks = 2020:2023) +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Cancer Expenditure by Diagnosis, Slovakia (absolute)",
       x = "Year", y = "Expenditure (million EUR €)", fill = "Diagnostic group") +
  theme_minimal(base_size = 11, base_family = "serif") +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 13, face = "bold"),
    legend.title    = element_text(size = 9),
    legend.text     = element_text(size = 7.5),
    legend.key.size = unit(0.4, "cm"),
    axis.title      = element_text(size = 11),
    panel.grid      = element_blank()
  )

print(slk_area)
#ggsave("slovakia_area_APPENDIX.png", slk_area,
#       width = 9, height = 5.5, dpi = 400, bg = "white")
```


*Figure 5. Estonia per component*

```{r}
#Estonia — per-component cancer expenditure by year 
#MAIN:     PPP$  (EUR / 0.583732)| APPENDIX: € millions

#MAIN TEXT — PPP$
est_ppp_cf <- 0.583732   # PPP conversion factor, Estonia 2024 (LCU per int'l $)

est_ppp_wide <- data.frame(
  Year        = 2020:2025,
  Medical     = c(209656068, 208724363, 193779454, 220426075, 243356243, 253085490),
  Pharmaceuticals = c(36572590, 39916309, 33261917, 35960324, 42146735, 40991489),
  Screening   = c(3610400, 5038765, 5621370, 6626341, 7511966, 8046626)
) %>%
  mutate(across(-Year, ~ .x / est_ppp_cf))   # EUR -> PPP$

est_ppp_long <- est_ppp_wide %>%
  pivot_longer(-Year, names_to = "Category", values_to = "Value") %>%
  group_by(Year) %>% mutate(pct = Value / sum(Value)) %>% ungroup() %>%
  mutate(Category = factor(Category,
           levels = c("Screening", "Pharmaceuticals", "Medical"),
           labels = c("Screening", "Pharmaceuticals", "Other medical services")))

est_ppp_pal <- c("Screening" = "#E69F00", "Pharmaceuticals" = "#0072B2",
                 "Other medical services" = "#009E73")

est_ppp_A <- ggplot(est_ppp_long, aes(x = factor(Year), y = pct, fill = Category)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.3,
           position = position_stack(reverse = TRUE)) +
  geom_text(aes(label = ifelse(pct >= 0.04,
                               paste0(round(pct*100), "%\n(",
                                      round(Value/1e6, 1), "M)"), "")),
            position = position_stack(vjust = 0.5, reverse = TRUE),
            colour = "white", size = 3.4, family = "serif", lineheight = 0.9) +
  scale_fill_manual(values = est_ppp_pal, name = NULL) +
  scale_y_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Per-Component Cancer Expenditure by Year, Estonia (PPP$)",
       x = NULL, y = "Share of total expenditure") +
  theme_classic(base_size = 15, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 16), legend.position = "bottom")

est_ppp_scr <- est_ppp_long %>% filter(Category == "Screening")

est_ppp_B <- ggplot(est_ppp_scr, aes(x = factor(Year), y = pct)) +
  geom_col(width = 0.72, fill = est_ppp_pal["Screening"]) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%\n(",
                               round(Value/1e6, 1), "M PPP$)")),
            vjust = -0.3, size = 3.4, family = "serif", lineheight = 0.9) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.036),
                     expand = expansion(mult = c(0, 0))) +
  labs(title = "Screening share (detail)", x = NULL, y = "Screening share") +
  theme_classic(base_size = 15, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 15))

est_ppp_combo <- est_ppp_A / est_ppp_B + plot_layout(heights = c(2.4, 1))
print(est_ppp_combo)

#ggsave("estonia_composition_PPP_MAIN.png", est_ppp_combo,
#       width = 9, height = 8, dpi = 400, bg = "white")

# APPENDIX — € millions
est_wide <- data.frame(
  Year        = 2020:2025,
  Medical     = c(209656068, 208724363, 193779454, 220426075, 243356243, 253085490),
  Pharmaceuticals = c(36572590, 39916309, 33261917, 35960324, 42146735, 40991489),
  Screening   = c(3610400, 5038765, 5621370, 6626341, 7511966, 8046626)
)

est_long <- est_wide %>%
  pivot_longer(-Year, names_to = "Category", values_to = "Value") %>%
  group_by(Year) %>% mutate(pct = Value / sum(Value)) %>% ungroup() %>%
  mutate(Category = factor(Category,
           levels = c("Screening", "Pharmaceuticals", "Medical"),
           labels = c("Screening", "Pharmaceuticals", "Other medical services")))

est_pal <- c("Screening" = "#E69F00", "Pharmaceuticals" = "#0072B2",
             "Other medical services" = "#009E73")

estA <- ggplot(est_long, aes(x = factor(Year), y = pct, fill = Category)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.3,
           position = position_stack(reverse = TRUE)) +
  geom_text(aes(label = ifelse(pct >= 0.04,
                               paste0(round(pct*100), "%\n(",
                                      round(Value/1e6, 1), "M\u20AC)"), "")),
            position = position_stack(vjust = 0.5, reverse = TRUE),
            colour = "white", size = 3.4, family = "serif", lineheight = 0.9) +
  scale_fill_manual(values = est_pal, name = NULL) +
  scale_y_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Per-Component Cancer Expenditure by Year, Estonia",
       x = NULL, y = "Share of total expenditure") +
  theme_classic(base_size = 15, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 16), legend.position = "bottom")

est_scr <- est_long %>% filter(Category == "Screening")

estB <- ggplot(est_scr, aes(x = factor(Year), y = pct)) +
  geom_col(width = 0.72, fill = est_pal["Screening"]) +
  geom_text(aes(label = paste0(round(pct*100, 1), "%\n(",
                               round(Value/1e6, 1), "M\u20AC)")),
            vjust = -0.3, size = 3.4, family = "serif", lineheight = 0.9) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.036),
                     expand = expansion(mult = c(0, 0))) +
  labs(title = "Screening share (detail)", x = NULL, y = "Screening share") +
  theme_classic(base_size = 15, base_family = "serif") +
  theme(plot.title = element_text(hjust = .5, size = 15))

est_combo <- estA / estB + plot_layout(heights = c(2.4, 1))
print(est_combo)
#ggsave("estonia_composition_EUR_APPENDIX.png", est_combo,
#       width = 9, height = 8, dpi = 400, bg = "white")
```

