---
title: "Turkmenistan Modelling Corr UPD"
output: html_document
date: "2026-07-15"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)   # print full numbers, no scientific notation
```


**Modelling the Cancer Expenditure in Turkmenistan**


*General Background Information about Turkmenistan*

Turkmenistan lacks a national insurance system, and most cancer financing comes from the government (which can also be a foreign loan or aid) and out-of-pocket (OOP) payments. Here, we model only the governmental expenditure that also likely includes funds of foreign origin.

Unlike in the other 17 countries, Turkmenistan does not publish an official cancer expenditure figure. However, its government/compulsory health financing is reported in the WHO Global Health Expenditure Database (GHED), which we use directly as the health-expenditure base instead of reconstructing it from the state budget.


*Formula for Estimating the Governmental Cancer Expenditure*

Formula to estimate Turkmenistan's cancer expenditure is:

Tur_cancer = Tur_prop * Tur_total

  + Tur_cancer - cancer expenditure in Turkmenistan in 2024
  + Tur_total - total public (government/compulsory) healthcare expenditure for Turkmenistan in 2024
  + Tur_prop - proportion of public health expenditure on cancer in 2024


*Estimating the Total Public Health Expenditure (Tur_total)*

We take Turkmenistan's government/compulsory health financing directly from GHED [1], using the HF.1 aggregate — "Government schemes and compulsory contributory health care financing schemes" — in national currency (million manat). This matches the definition of the public/compulsory base (GCHCFS) used for the other countries. The latest observed year in GHED is 2023, at 2,597 million manat. As Turkmenistan has no insurance system, HF.1 (2,597) and GGHE-D (2,568) differ by roughly 1%; we use HF.1 for consistency with the anchor countries.

We project the 2023 value to 2024 using Turkmenistan's GDP deflator (linked series, annual % change) [2], applying only the post-base-year rate (2024 = 6.83%). 

``` {r}
Tur_HF1_2023 <- 2597000000        #manat, GHED HF.1,latest observed 2023 [1]
defl_2024    <- 6.826100887/100 #GDP deflator, linked series, 2024 [2]

Tur_total <- Tur_HF1_2023 * (1 + defl_2024)   #2023->2024 projection

cat("GHED HF.1 base (2023), manat:", Tur_HF1_2023, "\n")
cat("2024 deflator applied (%):", round(defl_2024 * 100, 2), "\n")
cat("Estimated total public (government/compulsory) health expenditure, 2024, manat:",
    round(Tur_total), "\n")
cat("  same value in million manat:", round(Tur_total / 1e6), "\n")
```

*Estimating the Proportion of Health Expenditure Spent on Cancer (Tur_prop)*

1. We included 7 countries, where we could reliably identify what % of the governmental and/or insurance healthcare expenditure goes to cancer. Then, we use these countries to estimate the % of healthcare expenditure in Turkmenistan going to cancer.

``` {r}

# Kazakhstan, Currency: Tenge
Kaz_total <- 2227988999600 #[3]
Kaz_cancer <- 193938343000 #[4]
Kaz_prop <- Kaz_cancer / Kaz_total * 100

# Armenia, Currency: Dram
Arm_total <- 157092750466.177 #[5]
Arm_cancer <- 12959959308.6761 #[5]
Arm_prop <- Arm_cancer / Arm_total * 100

# Estonia, Currency: Euro
Est_total <- 2627749438.2 #[6][7]
Est_cancer <- 316782803 #[7]
Est_prop <- Est_cancer / Est_total * 100

# Russia, Currency: Ruble
Rus_total <- 6700000000000 #[8]
Rus_cancer <- 386649513200 #[9]
Rus_prop <- Rus_cancer / Rus_total * 100

# Slovenia, Currency: Euro
Slo_total <- 5352300000 #[10]
Slo_cancer <- 471756569.19 #[11]
Slo_prop <- Slo_cancer / Slo_total * 100

# Uzbekistan, Currency: Som
Uzb_total <- 4585752255600 #[13]
Uzb_cancer <- 164300000000 #[14]
Uzb_prop <- Uzb_cancer / Uzb_total * 100

# Azerbaijan, Currency: Manat
Aze_total <- 1829072000 #[15]
Aze_cancer <- 48300000 #[15]
Aze_prop <- Aze_cancer / Aze_total * 100


cat("% of public healthcare expenditure going to cancer in Kazakhstan is",
    round(Kaz_prop, 2), "%\n")

cat("% of public healthcare expenditure going to cancer in Armenia is",
    round(Arm_prop, 2), "%\n")

cat("% of public healthcare expenditure going to cancer in Estonia is",
    round(Est_prop, 2), "%\n")

cat("% of public healthcare expenditure going to cancer in Russia is",
    round(Rus_prop, 2), "%\n")

cat("% of public healthcare expenditure going to cancer in Slovenia is",
    round(Slo_prop, 2), "%\n")

cat("% of public healthcare expenditure going to cancer in Uzbekistan is",
    round(Uzb_prop, 2), "%\n")

cat("% of public healthcare expenditure going to cancer in Azerbaijan is",
    round(Aze_prop, 2), "%\n")

```

2. We assemble the anchor shares and weight them by how close each country's GDP per capita is to Turkmenistan's. The anchor shares are generally lowest for the lowest-income countries and highest for the highest-income country (Estonia), so cancer's share of public health spending appears to rise with national income. Weights are the inverse of the distance in GDP per capita [16], normalised to sum to one, so income-similar countries count most.

```{r}
tur_gdp_pc <- 5904.25773   #Turkmenistan GDP per capita, USD [16]

anchor_prop <- c(Kaz = Kaz_prop, Arm = Arm_prop, Est = Est_prop, Rus = Rus_prop, Slo = Slo_prop, Uzb = Uzb_prop, Aze = Aze_prop)

# GDP per capita (current USD), 2024 [16]
gdp_pc <- c(Kaz = 14154.6325, Arm = 8556.21407, Est = 31428.3548, Rus = 14961.89, Slo = 34301.0319, Uzb = 3337.45492, Aze = 7294.64276)

dist    <- abs(gdp_pc - tur_gdp_pc)
weights <- (1 / dist) / sum(1 / dist) #inverse-distance weights, sum to 1

cat("Anchor shares (%):\n");        print(round(anchor_prop, 2))
cat("\nGDP-per-capita weights:\n"); print(round(weights, 4))

prop_weighted <- sum(weights * anchor_prop)   # GDP-weighted central share
cat("\nGDP-weighted central share:", round(prop_weighted, 2), "%\n")
```


3. Now we apply the weighted share to Turkmenistan and report a central value with a plausible range:

```{r}
# GDP-weighted cancer expenditure
Tur_cancer_central <- Tur_total * prop_weighted / 100

# Range: lowermost and uppermost observed anchor shares applied to the same base
Tur_cancer_low  <- Tur_total * min(anchor_prop) / 100
Tur_cancer_high <- Tur_total * max(anchor_prop) / 100

cat("Public cancer expenditure, Turkmenistan 2024 (manat):\n")
cat("  central (GDP-weighted,", round(prop_weighted, 2), "%):",
    round(Tur_cancer_central), "\n")
cat("  lowermost  (", round(min(anchor_prop), 2), "% ):",
    round(Tur_cancer_low), "\n")
cat("  uppermost (", round(max(anchor_prop), 2), "%):",
    round(Tur_cancer_high), "\n\n")

cat("Public cancer expenditure, Turkmenistan 2024 (million manat):\n")
cat("  central   :", round(Tur_cancer_central / 1e6), "\n")
cat("  lowermost :", round(Tur_cancer_low  / 1e6), "\n")
cat("  uppermost :", round(Tur_cancer_high / 1e6), "\n")
```

We do not report a statistical confidence interval. The anchor number is small and not a random sample. Instead we give the central (GDP-weighted) estimate and a plausible range between the lowermost and uppermost anchor shares.




*References*

  + [1] Global Health Expenditure Database, World Health Organization. HF.1 "Government schemes and compulsory contributory health care financing schemes", Turkmenistan, Million NCU (latest observed year 2023). Available at: https://apps.who.int/nha/database/ViewData/Indicators/en. Accessed: 15.07.2026.
  + [2] GDP deflator: linked series (annual % change), Turkmenistan, 2024. World Bank (series NY.GDP.DEFL.KD.ZG.AD). Available at: https://data.worldbank.org/indicator/NY.GDP.DEFL.KD.ZG.AD?locations=TM. Accessed: 15.07.2026.
  + [3] Annual Report of the Mandatory Social Health Insurance Fund of Republic of Kazakhstan in 2024. Available at: https://msqory.kz/ru/eshche/o-fonde/godovye-otchety-fonda/. Accessed: 04.07.2026.
  + [4] Annual Report of Budget Implementation of the Ministry of Health of Republic of Kazakhstan in 2024. Available at: https://www.gov.kz/memleket/entities/dsm/documents/1?lang=ru&type=_14. Accessed: 04.07.2026.
  + [5] National Health Accounts of Armenia. 2024.
  + [6] European Observatory on Health Systems and Policies. Current health expenditure in 2024 reached historic high. Estonia. Available at: https://eurohealthobservatory.who.int/monitors/health-systems-monitor/updates/hspm/hspm-estonia-2023/current-health-expenditure-in-2024-reached-a-historic-high. Accessed: 05.07.2026.
  + [7] Annual Health Insurance Fund Report for 2024. Available at: https://tervisekassa.ee/en/organisation/annual-reports. Accessed: 04.07.2026.
  + [8] Annual Mandatory Social Health Insurance Report in 2024 of Russian federation. Available at: http://omsjournal.ru/wp-content/uploads/2025/08/oms_2025-4_web.pdf. Accessed: 04.07.2026.
  + [9] Minister of Health of Russia Announcement about Total Health Expenditure in 2024. Available at: https://tass.ru/ekonomika/20790859. Accessed: 04.07.2026.
  + [10] Health Expenditure and Source of Funding, 2024. Republic of Slovenia Statistical office. Available at: https://www.stat.si/StatWeb/en/news/Index/14355. Accessed: 04.07.2026.
  + [11] Cancer Expenditure provided to us by the Insurance Fund.
  + [12] GLOBOCAN 2024. International Agency for Research on Cancer. 2024.
  + [13] Periodic Financial Reporting of Ministry of Healthcare of Uzbekistan for the 4th quarter of 2024. Ministry of Healthcare of Uzbekistan. Available at: https://gov.uz/ru/ssv/sections/view/112959. Accessed: 07.07.2026.
  + [14] Ministry of Health of Uzbekistan, 2025 Healthcare Planning. https://gov.uz/uz/ssv/sections/view/54523
  + [15] Republic of Azerbaijan 2024 annual report on the implementation of the state budget. 2025. Accessed: 07.07.2026.
  + [16] GDP per capita (current US$). 2024. World Bank. Available at: https://data.worldbank.org/indicator/NY.GDP.PCAP.CD?name_desc=false. Accessed: 07.07.2026.